A systematic comparison of chloroplast genome assembly tools

### 2020-08-20 manuscript/bmc_article.tex
Preparation for revision submission

\newcommand*{\myexternaldocument}[1]{%
    \externaldocument{#1}%
    \addFileDependency{#1.tex}%
    \addFileDependency{#1.aux}%
}
\myexternaldocument{supplement}

@@

@@

\title{The landscape of chloroplast genome assembly tools}

@@

\author[
    addressref={aff1,aff3},
    %noteref={n1},
    email={jan.freudenthal@uni-wuerzburg.de}
    ]{\inits{JAF}\fnm{Jan A} \snm{Freudenthal}}
\author[
    addressref={aff1,aff2},
    email={simon-pfaff@gmx.de}
    ]{\inits{SP}\fnm{Simon} \snm{Pfaff}}
\author[
   addressref={aff1,aff3},
   email={niklas.terhoeven@uni-wuerzburg.de}
]{\inits{NT}\fnm{Niklas} \snm{Terhoeven}}
\author[
   addressref={aff1},
   email={arthur.korte@uni-wuerzburg.de}
]{\inits{AK}\fnm{Arthur} \snm{Korte}}
\author[
   addressref={aff1,aff3,aff4 },
   noteref={n2},
   email={markus.ankenbrand@uni-wuerzburg.de}
]{\inits{MJA}\fnm{Markus J} \snm{Ankenbrand}}
\author[
   addressref={aff1,aff2,aff5,aff6},                   % id's of addresses, e.g. {aff1,aff2}
   corref={aff6},                       % id of corresponding address, if any
   noteref={n2},                        % id's of article notes, if any
   email={frank.foerster@computational.bio.uni-giessen.de}   % email address
]{\inits{FF}\fnm{Frank} \snm{Förster}}

@@

\address[id=aff3]{%
  \orgname{AnaLife Data Science},
  \street{Wiesengrund 16},
  \postcode{97295 Waldbrunn}
  \city{W\"{u}rzburg},
  \cny{Germany}
}
\address[id=aff2]{%                           % unique id
  \orgname{Department of Bioinformatics, University of W\"{u}rzburg}, % university, etc
  \street{Biozentrum, Am Hubland},                     %
  \postcode{97074},                                % post or zip code
  \city{W\"{u}rzburg},                              % city
  \cny{Germany}                                    % country
}
\address[id=aff4]{%
  \orgname{Chair of Cellular and Molecular Imaging, Comprehensive Heart Failure Center, University Hospital W\"{u}rzburg},
  \street{Josef-Schneider-Str.\ 2},
  \postcode{97080}
  \city{W\"{u}rzburg},
  \cny{Germany}
}
\address[id=aff5]{%
  \orgname{Fraunhofer IME-BR},
  \street{Winchester Str.\ 2},
  \postcode{35394}
  \city{Gie\ss{}en},
  \cny{Germany}
}
\address[id=aff6]{%
  \orgname{Bioinformatics Core Facility of the University of Gie\ss{}en},
  \street{Heinrich-Buff-Ring 58},
  \postcode{35392}
  \city{Gie\ss{}en},
  \cny{Germany}
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Abstract begins here                 %%
%%                                          %%
%% Please refer to the Instructions for     %%
%% authors on http://www.biomedcentral.com  %%
%% and include the section headings         %%
%% accordingly for your article type.       %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstractbox}
\begin{abstract} % abstract
%One or two sentences providing a basic introduction to the field, comprehensible to a scientist in any discipline.
Chloroplasts are intracellular organelles that enable plants to conduct photosynthesis. They arose through the symbiotic integration of a prokaryotic cell into an eukaryotic host cell and still contain their own genomes with distinct genomic information.
Plastid genomes accommodate essential genes and are regularly utilized in biotechnology or phylogenetics.
%Two to three sentences of more detailed background, comprehensible to scientists in related disciplines.
Different assemblers that are able to asses the plastid genome, have been developed.
These assemblers often use data of whole genome sequencing experiments, which usually contain reads from the complete chloroplast genome.

%One sentence clearly stating the general problem being addressed by this particular study.
The performance of different assembly tools has never been systematically compared. %to each other concerning success rate or computational requirements.
%One sentence summarising the main result (with the words “here we show” or their equivalent).
Here we present a benchmark of seven chloroplast assembly tools, capable to succeed in more than \SI{60}{\percent} of known real data sets.
%Two or three sentences explaining what the main result reveals in direct comparison to what was thought to be the case previously, or how the main result adds to previous knowledge.
Our results show significant differences between the tested assemblers in terms of generating whole chloroplast genome sequences and computational requirements.
The examination of \num{105} data sets from species with unknown plastid genomes led to the assembly of \num{20} novel chloroplast genomes.
%\todo{on the novel set we only have 49/105 which is roughly 47 percent}
%One or two sentences to put the results into a more general context. 
We created docker images for each tested tool that are freely available for the scientific community and ensure reproducibility of the analyses.
These containers allow to analyze and screen data sets for chloroplast genomes using standard computational infrastructure.
%Two or three sentences to provide a broader perspective, readily comprehensible to a scientist in any discipline, may be included in the first paragraph if the editor considers that the accessibility of the paper is significantly enhanced by their inclusion. 
Thus, large scale screening for chloroplasts as hidden treasures within genomic sequencing data is feasible.

%\ff{Abstract}
%Useful as always: \url{https://cbs.umn.edu/sites/cbs.umn.edu/files/public/downloads/Annotated_Nature_abstract.pdf}
%The special issue will accept manuscripts presenting:
%
%- in-depth analyses of advantages
%
%-limitations of state-of-the-art methods
%
%-demonstration on a variety of representative datasets and simulations
%
%-clear guidelines/recommendations for the end-user
%
%-insight into future directions for methods
%\parttitle{First part title} %if any
%Text for this section.
%
%\parttitle{Second part title} %if any
%Text for this section.
\end{abstract}

@@

Structurally, two inverted repeats (\glspl{ir}) named \gls{ir}\textsubscript{A} and \gls{ir}\textsubscript{B} of \SIrange{10}{76}{\kilo\basepair} divide the chloroplast genome into a 
%long single copy (\gls{lsc}) and a short single copy (\gls{ssc}) region
\gls{lsc} and a \gls{ssc} region
\cite{palmer_1985}, which complicates genome assembly with short read technologies\cite{Wang2018}.
Moreover, the existence of different chloroplasts within a single individual, and thus multiple different chloroplast genomes, have been described for various plants \cite{corriveau_1988,Chat2002,Scar2016}. This phenomenon - called heteroplasmy - is only poorly understood in terms of its origin and evolutionary importance, but it impacts the assembly of whole chloroplast genomes.

@@

Different strategies have been developed to assemble chloroplast genomes \cite{twyford_strategies_2017}.
In general, obtaining a chloroplast genome from whole genome sequencing (WGS) data requires two steps: (1) extraction of chloroplast reads from the sequencing data;  
(2) assembly and resolution of the special circular structure including the \glspl{ir}.
The extraction of chloroplast reads can be achieved by mapping the reads to a reference chloroplast. \cite{Vinga2012}.
%\todo{gibt es dazu ein beispiel in unserem set? (niklas)}
A different approach that does not depend on the availability of a reference chloroplast, uses the higher coverage of reads originating from the chloroplast \cite{Chan2013}.
Here, a $k$-mer analysis can be used to extract the most frequent reads.
An example for this is implemented in \ce{} \cite{j_ankenbrand_chloroextractor:_2018}. %\todo{ce ist hier der einzig explizit genannte Assembler, das wirkt wie bias :) entfernen? Nope... Steht klar ein Beispiel und wir haben nun mal den angesehen... Würde ich jetzt einmal so drin lassen. Alternativ nehmen wir noch einen dazu}
%Arthur : Ich würds so lassen.
A third method, which is for example used by \np{} \cite{dierckxsens_novoplasty:_2017}, combines both approaches by using a reference chloroplast as seed and simultaneously assembling the reads based on $k$-mers.

%\ak{e.g. HomBlocks: A multiple-alignment construction pipeline for organelle phylogenomics based on locally collinear block searching, aber ist nicht inkluded, also können wir zitieren, müssen aber nicht}

@@

\section*{Results}
\subsection*{Performance metrics}
%\subsection*{Computational Resources}
%Table and figure of performance evaluation. Including RAM, I/O, runtime, parallelization ability, CPU usage. Free text description of results, range, mean/median, effect of input size, threads parameter, ...
All described tools have been tested with regard to their assembly time, memory and CPU utilization.

@@

Some assemblies failed to finish within our time limit of \SI{48}{\hour}. 
On average, the longest time to generate an assembly was taken by \ioga{} and \fp{} followed by \oa{} and \go{}.
The most time efficient tool was \ce{}, which was a little faster than \np{} and \cassp{}.
%% multithreading

Not all tested tools were able to benefit from having access to multiple threads.
Both \np{} and \oa{} required almost the same time independent of being allowed to utilize \numlist[list-final-separator={, or }]{1;2;4;8} threads.
In contrast, \cassp{}, \ce{}, \go  and \fp{} all profited from multi-threading settings  (\cref{fig:performance_runtime,fig:performance_memory_cpu,tab:perform25K_suppl,tab:perform250K_suppl,tab:perform2.5M_suppl}).

\subsubsection*{Memory and CPU Usage }
The peak and mean CPU usage, as well as peak memory and disk usage were recorded for all assemblers based on the same input data set and number of threads  (\cref{fig:performance_memory_cpu,tab:perform25K_suppl,tab:perform250K_suppl,tab:perform2.5M_suppl}).
In general, the size of the input data influenced the peak memory usage with the exception of \ce{} and \ioga{}.
Those two assemblers showed a memory usage pattern, which was less influenced by the size of the data.
The number of allowed threads had only a limited impact on the peak memory usage.
All programs profited from a higher number of threads, if the size of the input data was increased concerning their memory and CPU usage footprint.
%\todo{oben steht nicht alle? (niklas)}
% Frank: Hier geht es um Memory und CPU-Usage... Oben nur um Runtime
In contrast, the disk usage was independent of the size of the input data and the number of threads for all assemblers.

%but the factor between the programs exceeds a factor of \num{100}.

@@

Overall \go{} and \fp{} were the most successful tools on the simulated data while \cassp{} and \ioga{} were unable to successfully assemble any chloroplasts out of the \num{16} different data sets.
%The only assembler obtaining perfect results according to our score for the simulated data sets is \go{} (\cref{fig:simulated,tab:scores_simulated}).
%\ioga{} and \cassp{} showed the worst performance, being unable to fully assemble a single chloroplast out of \num{14} runs.
%\np{} performed second best with scores above \num{80} for all data sets, only failing to resolve the contigs into one single circular chloroplast assembly.
%The overall performance is best, when the input data consists purely of chloroplast reads.
%Only \ioga{} and \cassp{} failed to deliver any results under this scenario once.

\subsubsection*{Real data sets}
To evaluate the performance on real data, we used publicly available short read data from NCBI's SRA with existing reference chloroplasts.
We observed considerable differences for the tested assemblers, if we compared the generated alignments against the reference chloroplasts (\cref{fig:swarmplot}).
%\todo{hier ein halbsatz, was im score berücksichtigt wird? (niklas)}
%\todo{Jan: are these values correct? is table 3 up to date? This whole section might need to be rewritten, Yes all the tables and plots are up to date and to my knowledge correct}
The highest scores were achieved by \go{} with a median of \num{99.8} and \num{210} circular assemblies out of a total of 360 assemblies that resulted in an output (\cref{tab:scores_real}).
The performance of \go{} was followed by \fp{}, \np{}, \ioga{}, and \oa{}.
\fp{} outperformed \np{} and \oa{} in terms of score, producing twice as many \num{113} perfectly assembled chloroplast genomes (\np{} produced \num{58} and \oa{} \num{46} circular genomes).
\ioga{} and \cassp{} were both unable to assemble a circular, single-contig genome (\cref{tab:scores_real}, \cref{fig:upset}).
%\textit{swarm plot goes here}
%% Den upset plot eventuell erst in die Discussion reinbringen ? Da wir davon ableiten welche tools wir empfehlen?

\subsubsection*{Consistency}
Consistency was tested by re-running assemblies using the real data and comparison of the two assemblies (\cref{fig:consistency}).
\ce{} was the only tool able to reproduce the same scores in all runs (\cref{fig:consistency}).
%\todo{Frank or Jan: whoever wrote this sentence, what does it mean? \ce{} produced completely unsuccessful assemblies but consistently. Please re-write or remove! Dont think it was me}
\go{}, \oa{}, \cassp{}, and \ioga{} generated some assemblies that were unsuccessful in one run, but produced an output in the other attempt.
For these assemblers the scores were virtually identical if both runs were succesful. %\todo{Jan: identical? basically identical}.
Both \fp{} and \np{} show only minor changes for the successful assemblies, leading to arrow-shaped scatter plots (\cref{fig:consistency}).
%\ak{link to figure scatter plots is missing}
\ce{} appears to be the most robust assembler, showing no deviations between the two runs.

@@

For three assemblies, that were obtained by different assemblers, but not by \go{}, we kept one assembly obtained by \np{} and two from \fp{}.
All resulting \num{49} sequences have been annotated with GeSeq \cite{geseq}. 
The median number of distinct genes annotated were 80 for coding sequences, 4 for rRNA and 27 for tRNA (\cref{tab:feature_count_novel}, \cref{fig:feature_count_novel}). 
All sequences were stored in our repository \zenodorepo{}.
To avoid multi submissions of the same sequence to Genbank, all \num{49} sequences have been inspected against Genbank database via BLAST.
Finally, \num{20} sequences were uploaded to NCBI TPA:inferential (\cref{tab:accessions_suppl}) as novel chloroplast genomes.

@@

\subsection*{Guidelines for the end-user}
Given these results, our recommendation is to use \go{} as a default option for chloroplast assemblies. 
If \go{} does not produce a use-able assembly, \fp{} is a valid back-up solution that might be successful. 
This procedure maximizes the chance of effectively and efficiently recovering the circular chloroplast genome.
%\ak{what if both lead to qualitative different results?, always trust \go{} , or could one compute a high-confidence assembly by combining different results ?, might be a point for future directions.}
If both programs fail, it is recommended to try \np{} or manually fine-tune the parameters of the different tools.
It is obviously not possible to provide general guidelines, as the exact procedure will differ for different data sets.

@@

WGS data are also a rich source for chlorplast assemblies.
For nearly half of the analyzed data without available chloroplast genome, we could generate complete assemblies using at least one of the tools.
%\todo{Frank: where do the 60 percent come from?, is it not more like 40?}

Still, even with simulated (i.e. ``perfect'') data, not all tools succeeded in generating complete chloroplast assemblies.
Therefore, we determined the strengths and weaknesses of the specific tools and have provided guidelines for users.
It might however be necessary to combine different methods or manually explore the parameter space.
Ultimately, large scale studies reconstructing hundreds or thousands of chloroplast genomes are now feasible using the currently available tools. 
%Outlook: next step in the pipeline is annotation. There is a similar problem: many tools, which one to use? 
%A large scale study reconstructing chloroplasts throughout SRA (or \num{1001} genome project) is feasible and promising using the existing tools (estimated runtime, number of novel chloroplasts).
%A meta-tool that automatically combines the different programs could be developed.
%The benchmarking study could be augmented to automatically select the closest reference from the database and provide that to any tool that can take advantage of references.

\section*{Methods}
\subsection*{Data availability}
Source code for all methods used is available at \cite{github-benchmark-repo} and archived in zenodo under \zenodorepo{}.
All used assembly tools are hosted on GitHub (\cref{tab:toolsversions}) and are encapsulated in docker containers.
That docker containers are published on dockerhub \cite{dockerhub-benchmark} and are named with a leading \texttt{benchmark\_} (\cref{tab:dockerimages_suppl}).

To enable a fair comparison of all tools, we generated simulated sequencing data.
Those simulated data sets are stored at zenodo \zenododataset{}.
This study adheres to the guidelines for computational method benchmarking \cite{weber_essential_2018}.

@@

\subsection*{Our Setup}
We wanted to use a minimum of different parameter settings for all assembly programs to enable a fair comparison. 
Therefore, we decided to specify that all programs had to work based on two input files, representing the forward (\texttt{forward.fq}) and reverse  (\texttt{reverse.fq}) sequence file of a data set in FASTQ format.
Depending on the assembler, output files with different names and locations were generated.
Those different files were copied and renamed to ensure that each assembly approach produced the same output file (\texttt{output.fa}). 
Additionally, we set an environment variable for all programs to control the number of allowed threads.
All three requirements (defined input file names, defined output file name, thread number control via environment variable) were ensured by a simple wrapper script (\texttt{wrapper.sh}).
Finally, for a maximum of reproducibility, all programs were bundled into individual docker images based on a central base image which provides all the required software. Those docker images were used for the recording of the consumption of computational resources on a four Intel CPU-E7 8867 v3 system offering \SI{1}{\tera\byte} of RAM.
Furthermore, all our docker images have been converted into singularity containers for quantitative measurement on simulated and real data sets.
Singularity container were built from docker images for usage on an HPC-environment using Singularity v.2.5.2 \cite{kurtzer2017singularity}.
All singularity containers were run on %the Julia-HPC 
Intel® Xeon® Gold \num{6140} Processors using a Slurm workload manager version 17.11.8 \cite{Jette02slurm}. Assemblies were run on \num{4} threads using \SI{10}{\gibi\byte} RAM with a time limit of \SI{48}{\hour}.
\subsection*{Data}
\subsubsection*{Simulated data}
To avoid complications from sequencing errors and biological variation, we simulated perfect reads based on the \plant{At} (TAIR10) chloroplast and core genome assembly \cite{tair10}.
We used a sliding window approach with \texttt{seqkit} \cite{seqkit}. The exact commands are documented in \href{https://github.com/chloroExtractorTeam/benchmark/blob/master/03_representative_datasets.md}{\texttt{03\_representative\_datasets.md}} in \zenododataset{}.
For the final simulated data sets reads are based on the TAIR10 reference genome.
Different ratios between the \plant{At} core genome in combination with its mitochondrial sequence and the chloroplast sequence were generated (\num{0}:\num{1}, \num{1}:\num{10}, \num{1}:\num{100}, and \num{1}:\num{1000}).
The final data contained \SI{30}{\times} genome coverage and \SI{300}{\times} mitochondrial coverage, except the \num{0}:\num{1} ratio.
%\todo{Markus: mito ist richtig, oder?}
% Frank nach Rücksprache mit Markus: Ja... Ist Mitochondrium!
Additionally, we generated data with different read lengths (\SIlist{150;250}{\basepair}). We further sampled each data set to create another version containing exactly 2\,million read pairs.

%(circular for chloro and mito). No errors, no insert size variation, perfect quality, equal coverage.

@@

\subsubsection*{Novel data}
To evaluate the performance for chloroplasts without a reference in CpBase \cite{cpbase}, we sampled \num{105} data sets from the SRA \cite{sra2010} real data set described above (\cref{tab:novels_supp}).
For each entry within that novel data set the number of lineage splits between the source taxon and the related references from CpBase was calculated according to NCBI Taxonomy \cite{ncbitaxonomy}.
The final successful assembly of \num{49} new chloroplasts was manually inspected and rotated to follow the expected orientation and order of chloroplast genome parts.
Due to a lack of a clear definition, we followed the definition of \fp{} \cite{fastplast_orientation_issue}.
%\todo{Frank: Beschreibung des Submission prozesses und der Datenvorbereitung}
%to the closest relative as well as the last common ancestor have been calculated \todo{see table X (Frank)}. \todo{further description of the method (Frank)}

\subsection*{Evaluation Criteria}
\subsubsection*{Computational Resources}
%Due to computational resources are limited, we evaluated the effect of different input parameters.
We recorded the mean and the peak CPU usage, the peak memory consumption, and the size of the assembly folder for each program. 
As input data, we used different data sets comprising \numlist{25000;250000;2500000} read pairs sampled from  our simulated reads.
%Therefore, the simulated perfect read files have been read repeatedly to achieve the required number of read pairs.
We used our docker image setup (\cref{tab:dockerimages_suppl}) to run all assembly programs three times for each parameter setting.
The different settings combined different input data and different number of threads to use (\numlist{1;2;4;8}).

>>>>

\newcommand*{\myexternaldocument}[1]{%
    \externaldocument{#1}%
    \addFileDependency{#1.tex}%
    \addFileDependency{#1.aux}%
}
\myexternaldocument{additionalfile1}

@@

\newcommand{\crefsupp}[1]{Additional file\,1: \cref{#1}}

@@

\title{A systematic comparison of  chloroplast genome assembly tools}

@@

\author[
    addressref={aff1,aff3},
    %noteref={n1},
    email={jan.freudenthal@uni-wuerzburg.de}
    ]{\inits{JAF}\fnm{Jan A} \snm{Freudenthal}}
\author[
    addressref={aff1,aff2},
    email={simon-pfaff@gmx.de}
    ]{\inits{SP}\fnm{Simon} \snm{Pfaff}}
\author[
   addressref={aff1,aff3},
   email={niklas.terhoeven@uni-wuerzburg.de}
]{\inits{NT}\fnm{Niklas} \snm{Terhoeven}}
\author[
   addressref={aff1},
   email={arthur.korte@uni-wuerzburg.de}
]{\inits{AK}\fnm{Arthur} \snm{Korte}}
\author[
   addressref={aff1,aff3,aff4,aff5},
   noteref={n2},
   email={markus.ankenbrand@uni-wuerzburg.de}
]{\inits{MJA}\fnm{Markus J} \snm{Ankenbrand}}
\author[
   addressref={aff1,aff2,aff6,aff7,aff8},                   % id's of addresses, e.g. {aff1,aff2}
   corref={aff6},                       % id of corresponding address, if any
   noteref={n2},                        % id's of article notes, if any
   email={frank.foerster@computational.bio.uni-giessen.de}   % email address
]{\inits{FF}\fnm{Frank} \snm{Förster}}

@@

\address[id=aff3]{%
  \orgname{AnaLife Data Science},
  \street{Wiesengrund 16},
  \postcode{97295 Waldbrunn}
  \city{W\"{u}rzburg},
  \cny{Germany}
}
\address[id=aff2]{%                           % unique id
  \orgname{Department of Bioinformatics, University of W\"{u}rzburg}, % university, etc
  \street{Biozentrum, Am Hubland},                     %
  \postcode{97074},                                % post or zip code
  \city{W\"{u}rzburg},                              % city
  \cny{Germany}                                    % country
}
\address[id=aff4]{%
  \orgname{Chair of Cellular and Molecular Imaging, Comprehensive Heart Failure Center, University Hospital W\"{u}rzburg},
  \street{Josef-Schneider-Str.\ 2},
  \postcode{97080}
  \city{W\"{u}rzburg},
  \cny{Germany}
}
\address[id=aff5]{%
  \orgname{@IIMOG}
 }
\address[id=aff6]{%
  \orgname{Fraunhofer IME-BR},
  \street{Ohlebergsweg 12},
  \postcode{35392}
  \city{Gie\ss{}en},
  \cny{Germany}
}
\address[id=aff7]{%
  \orgname{Bioinformatics Core Facility of the University of Gie\ss{}en},
  \street{Heinrich-Buff-Ring 58},
  \postcode{35392}
  \city{Gie\ss{}en},
  \cny{Germany}
}
\address[id=aff8]{%
  \orgname{@frank\_foerster}
 }

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Abstract begins here                 %%
%%                                          %%
%% Please refer to the Instructions for     %%
%% authors on http://www.biomedcentral.com  %%
%% and include the section headings         %%
%% accordingly for your article type.       %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstractbox}
\begin{abstract} % abstract
\parttitle{Background} Chloroplasts are intracellular organelles that enable plants to conduct photosynthesis. They arose through the symbiotic integration of a prokaryotic cell into an eukaryotic host cell and still contain their own genomes with distinct genomic information. Plastid genomes accommodate essential genes and are regularly utilized in biotechnology or phylogenetics. Different assemblers that are able to assess the plastid genome have been developed. These assemblers often use data of whole genome sequencing experiments, which usually contain reads from the complete chloroplast genome.

\parttitle{Results} The performance of different assembly tools has never been systematically compared. Here we present a benchmark of seven chloroplast assembly tools, capable of succeeding in more than \SI{60}{\percent} of known real data sets. Our results show significant differences between the tested assemblers in terms of generating whole chloroplast genome sequences and computational requirements. The examination of \num{105}~data sets from species with unknown plastid genomes leads to the assembly of \num{20}~novel chloroplast genomes.

\parttitle{Conclusions} We create docker images for each tested tool that are freely available for the scientific community and ensure reproducibility of the analyses. These containers allow the analysis and screening of data sets for chloroplast genomes using standard computational infrastructure. Thus, large scale screening for chloroplasts within genomic sequencing data is feasible.
\end{abstract}

@@

Structurally, two inverted repeats (\glspl{ir}) named \gls{ir}\textsubscript{A} and \gls{ir}\textsubscript{B} of \SIrange{10}{76}{\kilo\basepair} divide the chloroplast genome into a 
\gls{lsc} and a \gls{ssc} region
\cite{palmer_1985}, which complicates genome assembly with short read technologies\cite{Wang2018}.
Moreover, the existence of different chloroplasts within a single individual, and thus multiple different chloroplast genomes, have been described for various plants \cite{corriveau_1988,Chat2002,Scar2016}. This phenomenon - called heteroplasmy - is only poorly understood in terms of its origin and evolutionary importance, but it impacts the assembly of whole chloroplast genomes.

@@

Different strategies have been developed to assemble chloroplast genomes \cite{twyford_strategies_2017}.
In general, obtaining a chloroplast genome from whole genome sequencing (WGS) data requires two steps: (1) extraction of chloroplast reads from the sequencing data;  
(2) assembly and resolution of the special circular structure including the \glspl{ir}.
The extraction of chloroplast reads can be achieved by mapping the reads to a reference chloroplast. \cite{Vinga2012}.
A different approach that does not depend on the availability of a reference chloroplast, uses the higher coverage of reads originating from the chloroplast \cite{Chan2013}.
Here, a $k$-mer analysis can be used to extract the most frequent reads.
An example for this is implemented in \ce{} \cite{j_ankenbrand_chloroextractor:_2018}. 
A third method, which is for example used by \np{} \cite{dierckxsens_novoplasty:_2017}, combines both approaches by using a reference chloroplast as seed and simultaneously assembling the reads based on $k$-mers.

@@

\section*{Results}
\subsection*{Performance metrics}
All described tools have been tested with regard to their assembly time, memory and CPU utilization.

@@

Some assemblies failed to finish within our time limit of \SI{48}{\hour}. 
On average, the longest time to generate an assembly was taken by \ioga{} and \fp{} followed by \oa{} and \go{}.
The most time efficient tool was \ce{}, which was a little faster than \np{} and \cassp{}.

Not all tested tools were able to benefit from having access to multiple threads.
Both \np{} and \oa{} required almost the same time independent of being allowed to utilize \numlist[list-final-separator={, or }]{1;2;4;8} threads.
In contrast, \cassp{}, \ce{}, \go  and \fp{} all profited from multi-threading settings  (\cref{fig:performance_runtime,fig:performance_memory_cpu} and \crefsupp{tab:perform25K_suppl,tab:perform250K_suppl,tab:perform2.5M_suppl}).

\subsubsection*{Memory and CPU Usage }
The peak and mean CPU usage, as well as peak memory and disk usage were recorded for all assemblers based on the same input data set and number of threads  (\cref{fig:performance_memory_cpu} and \crefsupp{tab:perform25K_suppl,tab:perform250K_suppl,tab:perform2.5M_suppl}).
In general, the size of the input data influenced the peak memory usage with the exception of \ce{} and \ioga{}.
Those two assemblers showed a memory usage pattern, which was less influenced by the size of the data.
The number of allowed threads had only a limited impact on the peak memory usage.
All programs profited from a higher number of threads, if the size of the input data was increased concerning their memory and CPU usage footprint.
In contrast, the disk usage was independent of the size of the input data and the number of threads for all assemblers.

@@

Overall \go{} and \fp{} were the most successful tools on the simulated data while \cassp{} and \ioga{} were unable to successfully assemble any chloroplasts out of the \num{16} different data sets.

\subsubsection*{Real data sets}
To evaluate the performance on real data, we used publicly available short read data from NCBI's SRA with existing reference chloroplasts.
We observed considerable differences for the tested assemblers, if we compared the generated alignments against the reference chloroplasts (\cref{fig:swarmplot}).
The highest scores were achieved by \go{} with a median of \num{99.8} and \num{210} circular assemblies out of a total of 360 assemblies that resulted in an output (\cref{tab:scores_real}).
The performance of \go{} was followed by \fp{}, \np{}, \ioga{}, and \oa{}.
\fp{} outperformed \np{} and \oa{} in terms of score, producing twice as many \num{113} perfectly assembled chloroplast genomes (\np{} produced \num{58} and \oa{} \num{46} circular genomes).
\ioga{} and \cassp{} were both unable to assemble a circular, single-contig genome (\cref{tab:scores_real}, \cref{fig:upset}).

\subsubsection*{Consistency}
Consistency was tested by re-running assemblies using the real data and comparison of the two assemblies (\cref{fig:consistency}).
\ce{} was the only tool able to reproduce the same scores in all runs (\cref{fig:consistency}).
\go{}, \oa{}, \cassp{}, and \ioga{} generated some assemblies that were unsuccessful in one run, but produced an output in the other attempt.
For these assemblers the scores were virtually identical if both runs were succesful.
Both \fp{} and \np{} show only minor changes for the successful assemblies, leading to arrow-shaped scatter plots (\cref{fig:consistency}).
\ce{} appears to be the most robust assembler, showing no deviations between the two runs.

@@

For three assemblies, that were obtained by different assemblers, but not by \go{}, we kept one assembly obtained by \np{} and two from \fp{}.
All resulting \num{49} sequences have been annotated with GeSeq \cite{geseq}. 
The median number of distinct genes annotated were 80 for coding sequences, 4 for rRNA and 27 for tRNA (\cref{tab:feature_count_novel}, \cref{fig:feature_count_novel}). 
All sequences were stored in our repository \zenodorepo{}.
To avoid multi submissions of the same sequence to Genbank, all \num{49} sequences have been inspected against Genbank database via BLAST.
Finally, \num{20} sequences were uploaded to NCBI TPA:inferential (\crefsupp{tab:accessions_suppl}) as novel chloroplast genomes.
Moreover, a search for the species name unveiled that \num{7} of the \num{20} sequences are used as ornamental plant, in folk medicine, or as crop plant.

@@

\subsection*{Guidelines for the end-user}
Given these results, our recommendation is to use \go{} as a default option for chloroplast assemblies. 
If \go{} does not produce a use-able assembly, \fp{} is a valid back-up solution that might be successful. 
This procedure maximizes the chance of effectively and efficiently recovering the circular chloroplast genome.
If both programs fail, it is recommended to try \np{} or manually fine-tune the parameters of the different tools.
It is obviously not possible to provide general guidelines, as the exact procedure will differ for different data sets.

@@

WGS data are also a rich source for chlorplast assemblies.
For nearly half of the analyzed data without available chloroplast genome, we could generate complete assemblies using at least one of the tools.

Still, even with simulated (i.e. ``perfect'') data, not all tools succeeded in generating complete chloroplast assemblies.
Therefore, we determined the strengths and weaknesses of the specific tools and have provided guidelines for users.
It might however be necessary to combine different methods or manually explore the parameter space.
Ultimately, large scale studies reconstructing hundreds or thousands of chloroplast genomes are now feasible using the currently available tools.

\section*{Methods}
\subsection*{Data availability}
Source code for all methods used is available at \cite{github-benchmark-repo} and archived in Zenodo under \zenodorepo{}.
All used assembly tools are hosted on GitHub (\cref{tab:toolsversions}) and are encapsulated in docker containers.
That docker containers are published on dockerhub \cite{dockerhub-benchmark} and are named with a leading \texttt{benchmark\_} (\crefsupp{tab:dockerimages_suppl}).

To enable a fair comparison of all tools, we generated simulated sequencing data.
Those simulated data sets are stored at Zenodo \zenododataset{}.
All resulting assemblies are available from Zenodo \cite{zenodoassemblies}.
This study adheres to the guidelines for computational method benchmarking \cite{weber_essential_2018}.

@@

\subsection*{Our Setup}
We wanted to use a minimum of different parameter settings for all assembly programs to enable a fair comparison. 
Therefore, we decided to specify that all programs had to work based on two input files, representing the forward (\texttt{forward.fq}) and reverse  (\texttt{reverse.fq}) sequence file of a data set in FASTQ format.
Depending on the assembler, output files with different names and locations were generated.
Those different files were copied and renamed to ensure that each assembly approach produced the same output file (\texttt{output.fa}). 
Additionally, we set an environment variable for all programs to control the number of allowed threads.
All three requirements (defined input file names, defined output file name, thread number control via environment variable) were ensured by a simple wrapper script (\texttt{wrapper.sh}).
Finally, for a maximum of reproducibility, all programs were bundled into individual docker images based on a central base image which provides all the required software. Those docker images were used for the recording of the consumption of computational resources on a four Intel CPU-E7 8867 v3 system offering \SI{1}{\tera\byte} of RAM.
Furthermore, all our docker images have been converted into singularity containers for quantitative measurement on simulated and real data sets.
Singularity container were built from docker images for usage on an HPC-environment using Singularity v.2.5.2 \cite{kurtzer2017singularity}.
All singularity containers were run on Intel® Xeon® Gold \num{6140} Processors using a Slurm workload manager version 17.11.8 \cite{Jette02slurm}. Assemblies were run on \num{4} threads using \SI{10}{\gibi\byte} RAM with a time limit of \SI{48}{\hour}.
\subsection*{Data}
\subsubsection*{Simulated data}
To avoid complications from sequencing errors and biological variation, we simulated perfect reads based on the \plant{At} (TAIR10) chloroplast and core genome assembly \cite{tair10}.
We used a sliding window approach with \texttt{seqkit} \cite{seqkit}. The exact commands are documented in \href{https://github.com/chloroExtractorTeam/benchmark/blob/master/03_representative_datasets.md}{\texttt{03\_representative\_datasets.md}} in \zenododataset{}.
For the final simulated data sets reads are based on the TAIR10 reference genome.
Different ratios between the \plant{At} core genome in combination with its mitochondrial sequence and the chloroplast sequence were generated (\num{0}:\num{1}, \num{1}:\num{10}, \num{1}:\num{100}, and \num{1}:\num{1000}).
The final data contained \SI{30}{\times} genome coverage and \SI{300}{\times} mitochondrial coverage, except the \num{0}:\num{1} ratio.
Additionally, we generated data with different read lengths (\SIlist{150;250}{\basepair}). We further sampled each data set to create another version containing exactly 2\,million read pairs.

@@

\subsubsection*{Novel data}
To evaluate the performance for chloroplasts without a reference in CpBase \cite{cpbase}, we sampled \num{105} data sets from the SRA \cite{sra2010} real data set described above (\crefsupp{tab:novels_supp}).
For each entry within that novel data set the number of lineage splits between the source taxon and the related references from CpBase was calculated according to NCBI Taxonomy \cite{ncbitaxonomy}.
The final successful assembly of \num{49} new chloroplasts was manually inspected and rotated to follow the expected orientation and order of chloroplast genome parts.
Due to a lack of a clear definition, we followed the definition of \fp{} \cite{fastplast_orientation_issue}.

\subsection*{Evaluation Criteria}
\subsubsection*{Computational Resources}
We recorded the mean and the peak CPU usage, the peak memory consumption, and the size of the assembly folder for each program. 
As input data, we used different data sets comprising \numlist{25000;250000;2500000} read pairs sampled from  our simulated reads.
We used our docker image setup (\crefsupp{tab:dockerimages_suppl}) to run all assembly programs three times for each parameter setting.
The different settings combined different input data and different number of threads to use (\numlist{1;2;4;8}).
